\pkg{Rgbp}: Bayesian Hierarchical Modeling and Frequentist Method Check

### 2013-04-17 jss/article.tex
print change

\subsection{Normal: 8 School Data}
Education Testing Service conducted randomized experiments in eight separate schools and obtained this dataset. It contains the coaching effects on SAT scores ($y_{j}, j=1, \ldots, 8$) and standard errors ($se_{j}, j=1, \ldots, 8$) of eight schools obtained after an analysis of covariance adjustment (Rubin, 1981).
\begin{CodeChunk}
\begin{CodeInput}
R> y  <- c(28,  8, -3,  7, -1,  1, 18, 12)
R> se <- c(15, 10, 16, 11,  9, 11, 10, 18)
\end{CodeInput}
\end{CodeChunk}

In the original paper, each school's coaching effect has approximately Normal sampling distribution with known sampling variance, \emph{i.e.}, standard error of each school is assumed to be known or to be accurately estimated. So, it is reasonable to think that each (group-level) coaching effect is distributed as independent Normal distribution given the unknown mean $\mu_{j}$ and known standard error:  $y_{j}\vert\mu_{j}\stackrel{ind}{\sim} \textrm{Normal}(\mu_{j}, se^{2}_{j}),~ j=1, \ldots, 8$. \pkg{Rgbp} includes this dataset and can be called by typing `\code{R> data(schools)}' on \proglang{R}.

\subsection{Poisson: 31 Hospital Data}
This dataset is about the medical profiling evaluations for Coronary Artery Bypass Graft (CABG) surgeries of 31 New York hospitals conducted in 1992 (Morris and Lysy, 2012). It comprises of the number of deaths within a month of CABG surgeries in each hospital ($z_{j},~j=1, \ldots, 31$) and total number of patients receiving CABG surgeries (case load) in each hospital ($n_{j},~j=1, \ldots, 31$). The below code is an example of input based on the last ten hospital data.
\begin{CodeChunk}
\begin{CodeInput}
R> z <- c( 14,   9,  15,  13,  35,  26,  25,  20,   35,   27)
R> n <- c(593, 602, 629, 636, 729, 849, 914, 940, 1193, 1340)
\end{CodeInput}
\end{CodeChunk}

@@

\subsection{Binomial: 18 Baseball Data}
This dataset contains information about batting averages of 18 major league baseball players through their first 45 official at-bats of the 1970 season (Efron and Morris, 1975). Also, it has two covariates, League and Position, showing in which league and in which position each player was playing. In this paper, we will use Position for a tutorial purpose. For convenience, we transform this variable into a binary indicator, which is 1 if a player was a outfielder and 0 otherwise. The code below shows a way to make inputs. If we have more than one covariate, for example, \code{x1} and \code{x2}, then `\code{R> x <- cbind(x1, x2)}' will be the right input of the \code{gbp} function.
\begin{CodeChunk}
\begin{CodeInput}
R> z <- c(18, 17, 16, 15, 14, 14, 13, 12, 11, 11, 10, 10, 10, 10, 10,  9,  8,  7)
R> n <- c(45, 45, 45, 45, 45, 45, 45, 45, 45, 45, 45, 45, 45, 45, 45, 45, 45, 45)
R> x <- c( 1,  1,  1,  1,  1,  0,  0,  0,  0,  1,  0,  0,  0,  1,  1,  0,  0,  0) 
\end{CodeInput}
\end{CodeChunk}
The data indicate that independent Binomial distribution is appropriate for each player's number of hits among 45 at-bats conditioning on the unknown true batting average $p_{j}$: $z_{j}\vert p_{j}\stackrel{ind}{\sim} \textrm{Binomial}(n_{j}, p_{j}), ~j=1, \ldots, 18$. This dataset is also a part of the package and can be called on \proglang{R} by `\code{R> data(baseball)}'.

@@

\subsection[Normal-Normal]{Normal-Normal}
\code{gbp} can construct a two-level Normal-Normal hierarchical model on the 8 school data. For reference,  $\sigma^{2}_{j}$ below is assumed to be known or to be accurately estimated, and subscript \emph{j} indicates \emph{j}-th school in the dataset.
\begin{equation}
y_{j}\vert \mu_{j} \stackrel{ind}{\sim}\textrm{Normal}(\mu_{j}, \sigma^{2}_{j}),
\end{equation}
\begin{equation}
\mu_{j}\vert \beta, A\stackrel{ind}{\sim}\textrm{Normal}(\mu_{0j}, A),
\end{equation}

@@

Here the adjustment for density maximization (Morris and Tang, 2011), called ADM, comes. It assumes the Beta distribution for shrinkage because this distribution is also defined between 0 and 1. In the end, ADM  estimates its posterior moments, \emph{i.e.}, $E(B_{j}\vert\textrm{data})$ and $Var(B_{j}\vert\textrm{data})$, without any trouble that MLE can cause. Please refer to Morris and Tang (2011) for more apparent advantages of ADM.
\\

@@

Once we are sure that the hierarchical modeling can be appropriate for our data, the following quetsion will be about the validity of interval estimates, the final product of this multilevel modeling. ``Does the 95\% (can be specified differently) confidence interval obtained via this Bayesian model-fitting process achieve 95\% confidence level  for any true parameter values?'' Our answer is ``yes'' and \pkg{Rgbp} has a function to assure this point. From now on, all the explanations will be based on the Binomial model.

@@

\subsection{Unbiased Coverage Probability}
Based on the generated pseudo-datasets, let's define an indicator variable, $I_{ij}$, which is 1 if $j$-th player's interval estimate from $i$-th pseudo-dataset includes $p_{ij}$ and 0 otherwise. One way to estimate the coverage probability is taking average over these indicator variables. We call it an unbiased coverage probability estimate. For example, $\bar{I}_{1}=\sum_{i=1}^{500}I_{i1}/500$ is the estimated coverage probability for the first player.

\subsection{Rao-Blackwellized Coverage Probability}
Based on the definition of Rao-Blackwellization, we define $E(I_{ij}\vert z_{ij}, r, \beta$), where $r$ and $\beta$ were given at first (see 5.1) and $z_{ij}$ is a sufficient statistic. This expectation is the same as Pr$(\hat{p}_{ij, low}\le p_{ij} \le\hat{p}_{ij, upp}\vert z_{ij}, r, \beta)$, where ($\hat{p}_{ij, low}$, $\hat{p}_{ij, upp}$) is $j$-th player's interval estimates on the $i$-th dataset. We can calculate this probability  because we know the distribution of $p_{ij} \vert z_{ij}, r, \beta$ in (9). Note that conditioning on $z_{ij}$ is equivalent to conditioning on $\mathbf{z}$ because we know $r$ and $\beta$. Then, we can estimate the first player's coverage probability by $\sum_{i=1}^{500}E(I_{i1}\vert z_{i1}, r, \beta)/500$, which is more accurate than the previous one.

@@

For this purpose, you gathered data of 31 hospitals in NY composed of the number of deaths within a month of CABG surgeries and total number of patients receiving CABG surgeries in each hospital. In addition, while you were looking for such information, suppose you could know that the  state-level death rate per exposure of this surgery in the past ten years was 0.02. 
\\

@@

As we can see in (6), the posterior mean, $(1-B_{j})y_{j} + B_{j}\lambda_{0}$, is a linear function of shrinkage, $B_{j}\equiv r / (r + n_{j})$, locating between the sample mean and prior mean ($\lambda_{0}=0.02$). It makes sense because $r$ can be interpreted as the amount of prior information and $n_{j}$ as the amount of observed information. If the second level has more information than the first level, then the sample mean shrinks towards the prior mean more than 50\%. This point is clear in the above output; as caseload increases, shrinkage decreases, depending less on the prior information.
\\

@@

  post.mode.alpha post.sd.alpha
1          -5.818         0.411
\end{CodeOutput}
\end{CodeChunk}
Since estimated $\alpha$ is -5.818, we can easily calculate $\hat{r}=\textrm{exp}(5.818)=336$, which will be helpful for understanding how each shrinkage was determined by comparing this value to the amount of observed information ($n_{j}$).
\\

>>>>

\subsection{Normal: 8 Schools Data}
Education Testing Service conducted randomized experiments in eight separate schools and obtained this dataset. It contains the coaching effects on SAT scores ($y_{j}, j=1, \ldots, 8$) and standard errors ($se_{j}, j=1, \ldots, 8$) of eight schools obtained after an analysis of covariance adjustment (Rubin, 1981).
\begin{CodeChunk}
\begin{CodeInput}
R> y  <- c(12, -3, 28,  7,  1,  8, 18, -1)
R> se <- c(18, 16, 15, 11, 11, 10, 10,  9)
\end{CodeInput}
\end{CodeChunk}

In the original paper, each school's coaching effect has approximately Normal sampling distribution with known sampling variance, \emph{i.e.}, standard error of each school is assumed to be known or to be accurately estimated. So, it is reasonable to think that each (group-level) coaching effect is distributed as an independent Normal distribution given the unknown mean $\mu_{j}$ and known standard error:  $y_{j}\vert\mu_{j}\stackrel{ind}{\sim} \textrm{Normal}(\mu_{j}, se^{2}_{j}),~ j=1, \ldots, 8$. \pkg{Rgbp} includes this dataset and can be called by typing `\code{R> data(schools)}' on \proglang{R}.

\subsection{Poisson: 31 Hospitals Data}
This dataset is about the medical profiling evaluations for Coronary Artery Bypass Graft (CABG) surgeries of 31 New York hospitals conducted in 1992 (Morris and Lysy, 2012). It comprises of the number of deaths within a month of CABG surgeries in each hospital ($z_{j},~j=1, \ldots, 31$) and total number of patients receiving CABG surgeries (case load) in each hospital ($n_{j},~j=1, \ldots, 31$). The below code is an example of input based on the last ten hospital data.
\begin{CodeChunk}
\begin{CodeInput}
R> z <- c( 14,   9,  15,  13,  35,  26,  25,  20,   35,   27)
R> n <- c(593, 602, 629, 636, 729, 849, 914, 940, 1193, 1340)
\end{CodeInput}
\end{CodeChunk}

@@

\subsection{Binomial: 18 Baseball Players Data}
This dataset contains information about batting averages of 18 major league baseball players through their first 45 official at-bats of the 1970 season (Efron and Morris, 1975). Also, it has one covariate, Position, showing in which position each player was playing. For convenience, we transform this variable into a binary indicator, which is 1 if a player was a outfielder and 0 otherwise. The code below shows a way to make inputs. If we have more than one covariate, for example, \code{x1} and \code{x2}, then `\code{R> x <- cbind(x1, x2)}' will be the right input of the \code{gbp} function.
\begin{CodeChunk}
\begin{CodeInput}
R> z <- c(18, 17, 16, 15, 14, 14, 13, 12, 11, 11, 10, 10, 10, 10, 10,  9,  8,  7)
R> n <- c(45, 45, 45, 45, 45, 45, 45, 45, 45, 45, 45, 45, 45, 45, 45, 45, 45, 45)
R> x <- c( 1,  1,  1,  1,  1,  0,  0,  0,  0,  1,  0,  0,  0,  1,  1,  0,  0,  0) 
\end{CodeInput}
\end{CodeChunk}
The data indicate that independent Binomial distribution is appropriate for each player's number of hits among 45 at-bats conditioning on the unknown true batting average $p_{j}$: $z_{j}\vert p_{j}\stackrel{ind}{\sim} \textrm{Binomial}(n_{j}, p_{j}), ~j=1, \ldots, 18$. This dataset is also a part of the package and can be called on \proglang{R} by `\code{R> data(baseball)}'.

@@

\subsection[Normal-Normal]{Normal-Normal}
\code{gbp} can construct a two-level Normal-Normal hierarchical model on the 8 school data. For reference,  $V_{j}~(\equiv \sigma^{2}/n_{j})$ below is assumed to be known or to be accurately estimated, and subscript \emph{j} indicates \emph{j}-th school in the dataset.
\begin{equation}
y_{j}\vert \mu_{j} \stackrel{ind}{\sim}\textrm{Normal}(\mu_{j}, V_{j}),
\end{equation}
\begin{equation}
\mu_{j}\vert \beta, A\stackrel{ind}{\sim}\textrm{Normal}(\mu_{0j}, A),
\end{equation}

@@

Here the adjustment for density maximization (Morris and Tang, 2011), called ADM, comes. It assumes the Beta distribution for shrinkage because this distribution is also defined between 0 and 1. In the end, ADM  estimates its posterior moments, \emph{i.e.}, $E(B_{j}\vert\textrm{data})$ and $Var(B_{j}\vert\textrm{data})$, without any trouble that MLE can cause. Refer to Morris and Tang (2011) for additional apparent advantages of ADM.
\\

@@

Once we are sure that the hierarchical modeling can be appropriate for our data, the following quetsion will be about the validity of interval estimates, the final product of this multilevel modeling. ``Does the 95\% (can be specified differently) confidence interval obtained via this Bayesian model-fitting process achieve 95\% confidence level  for any true parameter values?'' Our answer is ``yes'' and \pkg{Rgbp} has a function to check this point. From now on, all the explanations will be based on the Binomial model.

@@

\subsection{Simple Unbiased Estimates of Coverage Probabilities}
Based on the generated pseudo-datasets, let's define an indicator variable, $I_{ij}$, which is 1 if $j$-th player's interval estimate from $i$-th pseudo-dataset includes $p_{ij}$ and 0 otherwise. One way to estimate the coverage probability is taking average over these indicator variables. We call it an unbiased coverage probability estimate. For example, $\bar{I}_{1}=\sum_{i=1}^{500}I_{i1}/500$ is the estimated coverage probability for the first player.

\subsection{Rao-Blackwellized Unbiased estimates of Coverage Probabilities}
Based on the definition of Rao-Blackwellization, we define $E(I_{ij}\vert z_{ij}, r, \beta$), where $r$ and $\beta$ were given at first (see 5.1) and $z_{ij}$ is a sufficient statistic. This expectation is the same as Pr$(\hat{p}_{ij, low}\le p_{ij} \le\hat{p}_{ij, upp}\vert z_{ij}, r, \beta)$, where ($\hat{p}_{ij, low}$, $\hat{p}_{ij, upp}$) is $j$-th player's interval estimates on the $i$-th dataset. We can calculate this probability  because we know the distribution of $p_{ij} \vert z_{ij}, r, \beta$ in (9). Note that conditioning on $z_{ij}$ is equivalent to conditioning on $\mathbf{z}$ because we know $r$ and $\beta$. Then, we can estimate the first player's coverage probability by $\sum_{i=1}^{500}E(I_{i1}\vert z_{i1}, r, \beta)/500$, which is more accurate than the previous one.

@@

For this purpose, you gathered data of 31 hospitals in NY composed of the number of deaths within a month of CABG surgeries and total number of patients receiving CABG surgeries in each hospital. In addition, while you were looking for such information, suppose one knew that the state-level death rate per exposure of this surgery in the past ten years was 0.02. 
\\

@@

As we can see in (6), the posterior mean, $(1-B_{j})y_{j} + B_{j}\lambda_{0}$, is a linear function of shrinkage, $B_{j}\equiv r / (r + n_{j})$, locating between the sample mean and prior mean ($\lambda_{0}=0.02$). It makes sense because $r$ can be interpreted as the amount of prior information and $n_{j}$ as the amount of observed information. If the second level has more information than the first level, then the sample mean shrinks towards the prior mean more than 50\%. This point is clear in the above output; as caseload increases, shrinkage decreases, depending less on the prior distribution.
\\

@@

  post.mode.alpha post.sd.alpha
1          -5.818         0.411
\end{CodeOutput}
\end{CodeChunk}
Since estimated $\alpha$ is -5.818, we can easily calculate $\hat{r}=\textrm{exp}(5.818)=336$, which will be helpful for understanding how each shrinkage was determined by comparing this value to the observed sample sizes, $n_{j}$.
\\
